Automatic Parallelization and Transparent Fault Tolerance

### 2016-07-15 pubs/tfp16-draft/paper.tex
paper as submitted

\begin{abstract}

A claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Besides making large programs
easier to reason about, this enables automatic parallelization to various
degrees.  A second claimed trend in both scientific computing and functional
programming is the development of mechanisms for transparent fault tolerance.
Our project goal is the demonstration of a light-weight, higher-order,
polymorphic, pure functional language implementation in which we can
experiment with automatic parallelization strategies, varying degrees of
strictness, and mechanisms for transparent fault tolerance.
%
We do not consider speculative evaluation, or semantic strictness inferred by
program analysis, so potential parallelism is dictated by the specified degree of
strictness.

@@

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the USA and
other governments commit to a push for exascale computing ($10^{18}$ FLOPS in
a single system) within the next
decade~\cite{Exascale-roadmap,EUexascale,Exascale-org} a respected
member of the HPC community recently stated publicly that ``Fortran is
essential for exascale programming''~\cite{Heroux16}.  Existing US government
laboratory HPC codes run to hundreds of thousands of lines of Fortran each and
complete rewrites are infeasible.\footnote{The codes of our lab's British
  counterpart are similar.}  C++ has become more common for new code starts,
but their nature remains the same: huge codes with fair to non-existent
encapsulation of side effects, with multiple levels of parallelism sometimes
poorly abstracted, especially at the thread (shared-memory) and computational
accelerator (e.g.\ GPU) levels.  Thread-level parallel efficiency can be good
when the parallel model is bulk synchronous; tasking models are generally
avoided unless abstracted by, e.g., a C++ template or runtime library, because
of the programming complexity.

@@

\subsubsection{Rely on programmer discipline.}
Legacy Fortran scientific codes likely represent worst practices in state
encapsulation, with most data global and accessible to all; indeed, Fortran
was designed to make this easy via \texttt{COMMON} blocks---aggregates of
global variables---which further obfuscate program meaning by allowing
arbitrary naming of a block's variables at a subroutine/function granularity,
and requiring that entire blocks be brought into scope.  It is almost ironic,
then, that of the mainstream high-performance scientific programming languages
(Fortran, C, and C++), Fortran~95 was the first to introduce the \texttt{pure}
function qualifier, and with enforcement by the compiler.  Otherwise it is
entirely up to the programmer to make functions composable and thread safe.
%Here at LANL the use of pure functions (among other techniques for reducing
%global data dependencies) is being used to good effect to clean up some legacy
%code bases.

@@

\subsubsection{Use a pure functional language.}
Twenty years ago the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date, to no
effect whatsoever~\cite{Davis96}.\footnote{In 1996, on the LANL Thinking Machines
CM-5, the Top 1 supercomputer in 1993~\cite{top1:1993}.}
Today, however, there is a small number of laboratory HPC practitioners
who are developing an appreciation for pure functional programming in Haskell.
However, they see their needs as different than other Haskell users: performance
is paramount, array/vector
%~\cite{vector} 
is a (perhaps \emph{the}) primary data
structure, and non-strictness or laziness is more of a nuisance---even a
serious hindrance---than a convenience, much less ever algorithmically
essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
Johnsson's general attribute grammars~\cite{Johnsson87}, 
or other instances of \emph{tying the knot}~\cite{tying-the-knot}.

@@

\section{Project and Goals}
Given strict functional semantics wherein function arguments may be safely
evaluated before function evaluation or call, it is safe to evaluate the
arguments, and the function, in parallel, and that this applies recursively in
the (dynamic) expression tree (or graph), and similarly for data constructors
as functions.
%
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with varying degrees of
language strictness, and secondarily, with mechanisms for transparent fault
tolerance.  Light-weight has several implications.  First and foremost it
means that \emph{we are not attempting to compete with GHC in any
  respect}---succumbing to the temptation of recreating a significant subset
of GHC language, type system, or runtime features, or significant code
optimization, would simply divert us from our central goals.\footnote{We
  estimate that we will have, to within an order of magnitude, one person-hour
  of total available development time for each person-year that GHC currently
  represents, or in other units, one second for each hour.}
For another, it means that we are not undertaking significant morphing of GHC
itself: because this is in large part an undergraduate student project it must
be kept as simple as reasonably possible.
%
Our initial requirements are that
\begin{itemize}
\item The implementation must be feasible in terms of available effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item The generated code is C;
\item A certain consistency in tail calling in C must be achieved.
\end{itemize}
%
For feasibility we restrict ourselves to implicit parallelism on shared-memory
multiprocessors.  The requirement for tail calling is not primarily for
efficiency, rather for an efficient parallel implementation.  Achieving
this, and thereby sufficiently limiting the use of the native C stack, will be
discussed at some length.

@@

In short, our plan is to use GHC as our front-end to generate one of its
intermediate forms, STG or Core, then escape to our system from there.  There
are various possible ways to accomplish the this: using GHC's option to dump
STG or Core as text, hacking GHC itself, and, possibly, using GHC as a
library~\cite{ghc-as-library}.  Initially we started with our own STG.

@@

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG abstract machine.  Our partial description is based on that given
in \emph{EA}.

In STG the arity of a function is defined as the number of formal parameters
in the function definition, thus \texttt{id x = x} has by definition arity one
regardless of the type of the argument it might be applied to (e.g., some
function type).  We distinguish \emph{application}, e.g., \texttt{f x y},
which is a syntactic notion, from actual function \emph{call}. Thus \texttt{id
  id x} is an application of \texttt{id} to two arguments, but the underlying
\emph{FUN} (user- or system- defined function) when called is given exactly
one argument (the first one).

@@

\subsection{STG to C}
The back-end is written in Haskell and generates C code; the runtime system is
written in C/C++. This gives us the extensive optimization performed by modern
C compilers such as GCC and Clang/LLVM. The system fairly faithfully
implements the STG machine as described in \emph{EA} and dynamically
illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
transition rules which may be different than those for non-strict semantics.
Sharing is preserved.

@@

An almost unbounded amount of interesting development suggests itself; one
such path could in principle lead to the recreation of GHC\@.  However, given
the time constraints we must maintain sharp focus on the main goals:
implementing and evaluating mechanisms for automatic parallelism and fault
tolerance.  Currently we have a fully functioning serial implementation
including
\begin{itemize}
\item STG parser,
\item mini-Haskell parser, mini-Haskell to STG AST transformer,
\item partial coupling with the GHC front-end,
\item type inferencer,
\item various analysis and transformation passes,
\item code generator,
\item runtime system including a (serial) garbage collector and instrumentation framework,
\item continuous integration and testing system,
\item extensive test suite, automated testing.
\end{itemize}
Appendixes~\ref{app:minihaskell} and~\ref{app:ghclibrary}
describe the \emph{mini-Haskell} front-end and our approach to using GHC as a front end, respectively.
%

While wall-clock time can be used to simply compare the relative efficiency of
execution under various evaluation strategies, we seek more detailed
information.  We have in place a simple infrastructure for optionally logging
a thorough collection of run-time statistics such as number, size, and type of heap
and stack allocations, as well as various garbage collection statistics.  We also
measure cumulative time spent in various operations.  Future work includes processing
these measurement in an other than ad hoc fashion.

@@

For simplicity the implementation is 64-bit only.
%
The current, primitive parallel runtime is based on Pthreads and a
non-blocking work queue~\cite{Michael:1996}.  We are currently developing a
more realistic parallel runtime, experimenting with using the Argobots
user-level thread library~\cite{Seo:2015}.
%
%Bridging the gap between GHC STG and our STG, and the corresponding
%augmentation of the runtime is showing early signs of success.
%
An instrumentation infrastructure, integrated with the testing framework to
record and present time and space run-time statistics such as number of heap
and stack allocations and maximum stack depth, is planned for the near future.
%
The fault-tolerance mechanisms remain future work.
%
Parallel garbage collection is likely out of scope.
%
%We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
%small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. Marlow:2011 as well
%\cite{Harris:2005,Marlow:2008,Marlow:2009}

@@

%\emph{Nihil sub sole novum}---there is nothing new under the sun. % (Ecclesistes 1:9).
%every new idea has some sort of precedent or echo from the past~\cite{nihil}.
There is a vast body of related work; we highlight only the very most closely related.

@@

\subsubsection{Strict but still pure.}
\vspace{-0.1in}
%\paragraph{Strict but still pure.}
The concept of strict but still pure in some sense dates back to the concept
of applicative order reduction in the lambda calculus.  In terms of a
notionally concrete programming language the first thorough explications may
be Backus' FP~\cite{Backus:1978} (building on Landin's ``theory''~\cite{Landin64}).
% and follow-on FL implementation~\cite{AikenFL,FLreport89}.
%
A notable argument for the potential merits of a strict-by-default
pure functional language over a non-strict one is by Sheard, who also provided
a very simple embedded implementation~\cite{Sheard:2003}.
%
A plug-in for GHC (prior to 8.0) exists that makes function bodies strict in
their non-recursive \texttt{let} bindings using a transformation on GHC Core
that recursively expands those \texttt{let} bindings into \texttt{case}
bindings~\cite{Bolingbroke:2008,strict-ghc-plugin}.
%
The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order, functional language (or family of languages) compiler
and has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation, and the prospects
for its eventual completion are unknown~\cite{disciplined-disciple}.

@@

\vspace{-0.1in}
\subsubsection{GHC as a front-end.}
%\paragraph{GHC as a front-end.}
The Intel Labs Haskell Research Compiler (HRC) is the most closely related
work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
planned approach they use GHC as the front end, but exit GHC at the final GHC
Core stage, just before transformation to STG.  The essence of their effort is
to target an existing, highly optimizing, Intel functional language compiler
that is ``largely language agnostic'' and performs (among many other
optimizations) SIMD vectorization.

@@

\section{Mini-Haskell}
\label{app:minihaskell}
As mentioned, a current near-term goal is to use GHC as our front-end to
generate STG (or Core) as input to our system.  In the meantime, as a stop-gap
measure to make the generation of test cases easier, we have implemented a
\emph{mini-Haskell} front-end that supports a subset of standard Haskell
syntax with \emph{MagicHash} enabled, with a single extension.  Recall that
\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote
an unboxed literal value.  Our extension (as for STG) is the provision of
the optional \texttt{unboxed} keyword in datatype definitions to allow
user-defined unboxed data types, e.g.,
{\footnotesize
\begin{verbatim}
   data unboxed Bool# = False# | True# .
\end{verbatim}}

Writing STG in the concrete syntax conceived for our project is tedious and
prone to errors.  Compared to Haskell, the eventual source language for the
compiler, STG is relatively verbose, requiring explicit block and expression
delimiters, and classification of heap objects. Though its general structure
is similar to primitive Haskell, it lacks many of the conveniences that make
Haskell a pleasure to program in.  The motivation for a more expressive
language for testing our compiler and runtime system in their nascent stages
is then clear.  This language, mini-Haskell, was progressively ``grown'' from
the STG syntax towards a subset of Haskell itself, removing much of the
frustrating syntactic cruft and adding some of the core features that are
often taken for granted in Haskell (and other languages).  An alternative
approach would translate from mini-Haskell to GHC Core, then either to STG or,
in the case that the back end eventually take, compiling Core directly.

@@

\vspace{-0.1in}
\subsubsection{Fancy pattern matching.}
STG supports pattern matching but only in a simple form and only in
\texttt{case} expressions.  Mini-Haskell extends this to something approaching
what is found in Haskell.  Functions, including lambda expressions, may match on
all their arguments. Named functions may have multiple definitions to exhaust
the combinations of patterns in their arguments, and patterns in all contexts
(functions and case expressions) may be nested arbitrarily deep.  This feature
greatly simplifies the task of defining more complex behavior.  Without it,
each matched parameter of a function requires an additional level of nested
\texttt{case} expressions. The same is true of each level of nesting within a
single pattern.
%
Even relatively simple functions can be painful to write, such as \texttt{unzip}:
{\footnotesize
\begin{verbatim}
   unzip = FUN(list -> 
     case list of s1 { Nil -> let {res = CON( Pair nil nil)} in res;
                    Cons x xs -> case x of s2 {
                      Pair a b -> case unzip xs of s3 {
                        Pair as bs -> let {aas = CON(Cons a as);
                                           bbs = CON(Cons b bs);
                                           res = CON(Pair aas bbs)}
                                      in res }}})
\end{verbatim}}
\noindent In mini-Haskell this becomes
{\footnotesize
\begin{verbatim}
   unzip :: List (Pair a b) -> Pair (List a) (List b)
   unzip Nil = Pair Nil Nil
   unzip (Cons (Pair a b) xs) =
     case unzip xs of
       Pair as bs -> Pair (Cons a as) (Cons b bs)

\end{verbatim}}

Of course, this complex pattern matching must eventually be transformed into
the primitive, nested case expressions. Fortunately there are well-documented
algorithms for achieving this, and Philip Wadler's chapter
% in \emph{The Implementation of Functional Programming Languages}
on the compilation of this type of pattern matching proved
invaluable~\cite{Jones:1987}.
%Trimming Mini-Haskell to fit page limit
%\begin{comment} 
Wadler describes the \emph{match} function
which, in its simplest form, produces a correct, but potentially large
translation with some redundancy.  This function was implemented with only
minor modifications to fit the shape of mini-Haskell's abstract form.  The
implementation is augmented by some light flow-analysis to eliminate some of
the redundancy and make the resulting code easier to read.  There are further
optimizations suggested by Wadler to address the size of the generated code,
for example, when multiple branches lead to the same large expression, but the
additional complexity outweighs the benefit for a testing language.
%\end{comment}

\vspace{-0.1in}
\subsubsection{Less significant features.}
Mini-Haskell provides some smaller conveniences that don't warrant individual
sections but are worth mentioning.
\begin{itemize}
\item \emph{Constructors as functions.}
  As in Haskell, constructors are first class, acting as true functions.  This
  is implemented by simply generating a function that wraps the appropriate
  \texttt{let}-bound \texttt{CON} object in STG.  %% Any partially applied
  %% constructor generates a call to this function.  Fully applied constructors
  %% generate local \texttt{let}-bound objects to avoid an unnecessary function
  %% call.

@@

\item \emph{Type annotations.}  The ability to assert the type of some
  variable is often valuable as a sanity-check and provides documentation that
  the compiler can verify.  Mini-Haskell's type annotations (or signatures)
  behave like those in Haskell, being passed to the type checker as initial
  assumptions.

@@

It is a testament to the talents of the GHC contributors and
maintainers that we have managed to come even this far.  Their
documentation and exposed interfaces have made navigating and
comprehending the project internals much more pleasant than expected.

@@

There are several hundred small STG and mini-Haskell programs in the test
suite, which is partitioned into a number of directories which control what
testing is done. Given our goal of exploring various levels of strictness,
most tests are run with all evaluation strategies (three degrees of function
strictness by two degrees of constructor strictness).  Interesting classes of
test programs are those that require less-strict or sharing semantics.  For
these programs we test not only that the test leads to the expected result
with less strict semantics, but also that the program returns a
\emph{BLACKHOLE} in the case of a stricter semantics. Students who were
Haskell/STG beginners have contributed substantially to the test suite. We
have also found that the test suite has been of use as a learning tool for
these students to discover what levels of non-strictness are required for their
programs, a benefit that we did not initially envision. Over time we have
developed extensive STG and mini-Haskell preludes of common programming
patterns to aid in the writing of test programs.

Another important class of test programs are those that should fail with a
known error, e.g., a parsing or typing error.  For these programs the test
suite checks the output against an expected error regular expression. This
class of test programs has proved valuable in development and debugging of
both the STG and mini-Haskell front-ends.

@@

We have also implemented a simple logging facility which can be controlled at 
both compile and run time which helps in determining why a test is failing.

>>>>

\begin{abstract}
{\color{OliveGreen} % \bf
\emph{Draft paper notes, per the CfP:  Ringo, Anderson, and Marks are undergraduate
students.  This is a ``project'' paper.}
}

There is an increasing appreciation and practice of functional programming in
scientific computing, and in particular strict (by default) functional
programming, in various guises.  Besides making large programs easier to
reason about, this enables automatic parallelization to various degrees.  A
second trend in both scientific computing and functional programming is the
development of mechanisms for transparent fault tolerance.  Our project goal
is the demonstration of a light-weight, higher-order, polymorphic, pure
functional language implementation in which we can experiment with automatic
parallelization strategies, varying degrees of strictness, and mechanisms for
transparent fault tolerance.
%
We do not consider speculative evaluation, or semantic strictness inferred by
program analysis, so potential parallelism is dictated by the specified degree of
strictness.

@@

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as 
various governments commit to a push for exascale computing ($10^{18}$ FLOPS in
a single system) within the next
decade~\cite{Exascale-roadmap,EUexascale,Exascale-org} a respected
member of the HPC community recently stated publicly that ``Fortran is
essential for exascale programming,'' meaning that existing codes
will be adapted, not rewritten~\cite{Heroux16}.  Existing US government
laboratory HPC codes run to hundreds of thousands of lines of Fortran each and
complete rewrites are infeasible.\footnote{The codes of our lab's British
  counterpart are similar.}  C++ has become more common for new code starts,
but their nature remains the same: huge codes with fair to non-existent
encapsulation of side effects, with multiple levels of parallelism sometimes
poorly abstracted, especially at the thread (shared-memory) and computational
accelerator (e.g.\ GPU) levels.  Thread-level parallel efficiency can be good
when the parallel model is bulk synchronous; tasking models are generally
avoided unless abstracted by, e.g., a C++ template or runtime library, because
of the programming complexity.

@@

\subsubsection{Rely on programmer discipline.}
Legacy Fortran scientific codes likely represent worst practices in state
encapsulation, with most data global and accessible to all; indeed, Fortran
was designed to make this easy via \texttt{COMMON} blocks---aggregates of
global variables---which further obfuscate program meaning by allowing
arbitrary naming of a block's variables at a subroutine/function granularity,
and requiring that entire blocks be brought into scope.  It is almost ironic,
then, that of the mainstream high-performance scientific programming languages
(Fortran, C, and C++), Fortran~95 was the first to introduce the \texttt{pure}
function qualifier, and with enforcement by the compiler.  Otherwise it is
entirely up to the programmer to make functions composable and thread safe.
Here at LANL the concept of \emph{pure functional} is being used to good
effect to clean up---restructure, modularize, reduce to composable
components---some legacy code bases.
%Here at LANL the use of pure functions (among other techniques for reducing
%global data dependencies) is being used to good effect to clean up some legacy
%code bases.

@@

\subsubsection{Use a pure functional language.}
Twenty years ago the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date, to no
effect whatsoever~\cite{Davis96}.\footnote{In 1996 on the LANL Thinking Machines
CM-5, the Top 1 supercomputer in 1993~\cite{top1:1993}.}
Today, however, there is a small number of laboratory HPC practitioners
who are developing an appreciation for pure functional programming in Haskell.
However, they see their needs as different than other Haskell users: performance
is paramount, array/vector
%~\cite{vector} 
is a (perhaps \emph{the}) primary data
structure, and non-strictness or laziness is more of a nuisance---even a
serious hindrance---than a convenience, much less ever algorithmically
essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
Johnsson's general attribute grammars~\cite{Johnsson87}, 
or other instances of \emph{tying the knot}~\cite{tying-the-knot}.

@@

\section{Project and Goals}
Given strict functional semantics wherein function arguments may be safely
evaluated before function evaluation or call, it is safe to evaluate the
arguments, and the function, in parallel, and that this applies recursively in
the (dynamic) expression tree (or graph), and similarly for data constructors
as functions.
%
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with varying degrees of
language strictness, and secondarily, with mechanisms for transparent fault
tolerance.  Light-weight has several implications.  First and foremost it
means that \emph{we are not attempting to compete with GHC in any
  respect}---succumbing to the temptation of recreating a significant subset
of GHC language, type system, or runtime features, or significant code
optimization, would simply divert us from our central goals.\footnote{We
  estimate that we will have, to within an order of magnitude, one person-hour
  of total available development time for each person-year that GHC currently
  represents, or in other units, one second for each hour.}
For another, it means that we are not undertaking morphing of GHC
itself: the system is huge and complex.
%
Our initial requirements are that
\begin{itemize}
\item The implementation must be feasible in terms of available effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item The generated code is C;
\item A certain consistency in tail calling in C must be achieved.
\end{itemize}
%
For feasibility we restrict ourselves to implicit parallelism on shared-memory
multiprocessors.  The requirement for tail calling is not primarily for
efficiency, rather for an efficient parallel implementation.  Achieving
this, and thereby sufficiently limiting the use of the native C stack, will be
discussed at some length.

@@

In short, we will use GHC as the front-end to generate one of its
intermediate forms, STG or Core, then escape to our system from there.  There
are various possible ways to accomplish the this: using GHC's option to dump
STG or Core as text, hacking GHC itself, or as we have chosen, using GHC as a
library~\cite{ghc-as-library}.  Initially we started with our own STG.

@@

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG abstract machine.  Our partial description is based on that given
in \emph{EA}.
%
In STG the arity of a function is defined as the number of formal parameters
in the function definition, thus \texttt{id x = x} has by definition arity one
regardless of the type of the argument it might be applied to (e.g., some
function type).  We distinguish \emph{application}, e.g., \texttt{f x y},
which is a syntactic notion, from actual function \emph{call}. Thus \texttt{id
  id x} is an application of \texttt{id} to two arguments, but the underlying
\emph{FUN} (user- or system- defined function) when called is given exactly
one argument (the first one).

@@

\subsection{STG to C}
The back-end is written in Haskell and generates C code; the runtime system is
written in C/C++. This gives us the extensive optimization performed by modern
C compilers such as GCC and Clang/LLVM. The system fairly faithfully
implements the STG machine as described in \emph{EA} and dynamically
illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
transition rules which may be different than those for non-strict semantics.

@@

An almost unbounded amount of interesting development suggests itself; one
such path could in principle lead to the recreation of GHC\@.  However, given
the time constraints we must maintain sharp focus on the main goals:
implementing and evaluating mechanisms for automatic parallelism and fault
tolerance.  Currently we have a fully functioning serial implementation
including
\begin{itemize}
\item STG parser,
\item mini-Haskell parser, mini-Haskell to STG AST transformer,
\item partial coupling with the GHC front-end,
\item type inferencer,
\item various analysis and transformation passes,
\item code generator,
\item runtime system including a (serial) garbage collector and instrumentation framework, and
\item continuous integration and testing system, extensive test suite.
\end{itemize}
Appendixes~\ref{app:minihaskell} and~\ref{app:ghclibrary} describe the
mini-Haskell front-end and our approach to using GHC as a front end,
respectively.
%

While wall-clock time can be used to simply compare the relative efficiency of
execution under various evaluation strategies, we seek more detailed
information.  We have in place a simple infrastructure for optionally logging
a thorough collection of run-time statistics such as number, size, and type of
heap and stack allocations, as well as various garbage collection statistics.
We also measure cumulative time spent in various operations.  Future work
includes integrating this with the testing system, processing these statistics
in an other than ad hoc fashion.

@@

For simplicity the implementation is 64-bit only.
%
The current, primitive parallel runtime is based on Pthreads and a
non-blocking work queue~\cite{Michael:1996}.  We are currently developing a
more realistic parallel runtime, experimenting with using the Argobots
user-level thread library~\cite{Seo:2015}.
%
%Bridging the gap between GHC STG and our STG, and the corresponding
%augmentation of the runtime is showing early signs of success.
%
The fault-tolerance mechanisms remain future work.
%
Parallel garbage collection is likely out of scope.
%
%We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
%small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. Marlow:2011 as well
%\cite{Harris:2005,Marlow:2008,Marlow:2009}

@@

%\emph{Nihil sub sole novum}---there is nothing new under the sun. % (Ecclesistes 1:9).
%every new idea has some sort of precedent or echo from the past~\cite{nihil}.
There is a vast body of related work; we highlight only the most similar to ours.

@@

\subsubsection{Strict but still pure.}
\vspace{-0.1in}
%\paragraph{Strict but still pure.}
The concept of strict but still pure in some sense dates back to the concept
of applicative order reduction in the lambda calculus.  In terms of a
notionally concrete programming language the first thorough explications may
be Backus' FP~\cite{Backus:1978} (building on Landin's ``theory''~\cite{Landin64}).

A notable argument for the potential merits of a strict-by-default
pure functional language over a non-strict one is by Sheard, who also provided
a very simple embedded implementation~\cite{Sheard:2003}.

A plug-in for GHC (prior to 8.0) exists that makes function bodies strict in
their non-recursive \texttt{let} bindings using a transformation on GHC Core
that recursively expands those \texttt{let} bindings into \texttt{case}
bindings~\cite{Bolingbroke:2008,strict-ghc-plugin}.

Manticore is a derivative of Standard ML with no mutable state and both
implicit and explicit (threaded) mechanisms for parallelism.  Threads
communicate and synchronize via synchronous channels~\cite{Manticore}.
Numerous other ML derivatives, e.g.\ MultiMLton~\cite{MultiMLton},
OCaml~\cite{OCaml}, Poly/ML~\cite{PolyML}, and more distantly
NESL~\cite{NESL}, support various forms of parallelism.

The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order, functional language (or family of languages) compiler
and has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation~\cite{disciplined-disciple}.

@@

Futhark is a statically-typed, data-parallel,
pure functional language targeting GPUs via OpenCL~\cite{Futhark}.  Notably, it
supports regular nested data-parallelism, and a purity-preserving
update-in-place mechanism via a uniqueness type system.

\vspace{-0.1in}
\subsubsection{GHC as a front-end.}
%\paragraph{GHC as a front-end.}
The Intel Labs Haskell Research Compiler is the most closely related work to ours
of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our approach they
use GHC as the front end, but exit GHC at the final GHC Core stage, just
before transformation to STG.  The essence of their effort is to target an
existing, highly optimizing, Intel functional language compiler that is
``largely language agnostic'' and performs (among many other optimizations)
SIMD vectorization.

@@

\section{Mini-Haskell}
\label{app:minihaskell}
As mentioned, we are transitioning to GHC as our front-end to
generate STG (or Core) as input to our system.  Previously, as a stop-gap
measure to make the generation of test cases easier, we implemented a
\emph{mini-Haskell} front-end that supports a subset of standard Haskell
syntax with \emph{MagicHash} enabled, with a single extension.  Recall that
\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote
an unboxed literal value.  Our extension (as for STG) is the provision of
the optional \texttt{unboxed} keyword in datatype definitions to allow
user-defined unboxed data types, e.g.,
{\footnotesize
\begin{verbatim}
   data unboxed Bool# = False# | True# .
\end{verbatim}}

Writing STG in the concrete syntax conceived for our project is tedious and
prone to errors.  Compared to Haskell, the eventual source language for the
compiler, STG is relatively verbose, requiring explicit block and expression
delimiters, and classification of heap objects. Though its general structure
is similar to primitive Haskell, it lacks many of the conveniences that make
Haskell convenient to program in.  The motivation for a more expressive
language for testing our compiler and runtime system in their nascent stages
is then clear.  This language, mini-Haskell, was progressively ``grown'' from
the STG syntax towards a subset of Haskell itself, removing much of the
frustrating syntactic cruft and adding some of the core features that are
often taken for granted in Haskell (and other languages).  An alternative
approach would translate from mini-Haskell to GHC Core, then either to STG or,
in the case that the back end may eventually take, compiling Core directly.

@@

\vspace{-0.1in}
\subsubsection{Nested pattern matching.}
STG supports pattern matching but only in a simple form and only in
\texttt{case} expressions.  Mini-Haskell extends this to something approaching
what is found in Haskell.  Functions, including lambda expressions, may match on
all their arguments. Named functions may have multiple definitions to exhaust
the combinations of patterns in their arguments, and patterns in all contexts
(functions and case expressions) may be nested arbitrarily deep.  This feature
greatly simplifies the task of defining more complex behavior.  Without it,
each matched parameter of a function requires an additional level of nested
\texttt{case} expressions. The same is true of each level of nesting within a
single pattern.
%
Even relatively simple functions can be painful to write, such as \texttt{unzip}:
{\footnotesize
\begin{verbatim}
   unzip = FUN(list -> 
     case list of s1 { Nil -> let {res = CON( Pair nil nil)} in res;
                    Cons x xs -> case x of s2 {
                      Pair a b -> case unzip xs of s3 {
                        Pair as bs -> let {aas = CON(Cons a as);
                                           bbs = CON(Cons b bs);
                                           res = CON(Pair aas bbs)}
                                      in res }}})
\end{verbatim}}
\noindent In mini-Haskell this becomes
{\footnotesize
\begin{verbatim}
   unzip :: List (Pair a b) -> Pair (List a) (List b)
   unzip Nil = Pair Nil Nil
   unzip (Cons (Pair a b) xs) =
     case unzip xs of
       Pair as bs -> Pair (Cons a as) (Cons b bs)

\end{verbatim}}
This nested pattern matching must eventually be transformed into
primitive, nested case expressions.  There are well-documented
algorithms for achieving this, and Wadler's description
% in \emph{The Implementation of Functional Programming Languages}
of the compilation of this type of pattern matching proved
invaluable~\cite{Jones:1987}.
%Trimming Mini-Haskell to fit page limit
%\begin{comment} 
Wadler describes the \emph{match} function which, in its simplest form,
produces a correct but potentially large translation with some redundancy.
This function was implemented with only minor modifications to fit the shape
of mini-Haskell's abstract form.  The implementation is augmented by some
light flow-analysis to eliminate some of the redundancy and make the resulting
code easier to read.  Wadler suggests further optimizations to address the
size of the generated code, for example, when multiple branches lead to the
same large expression, but the additional complexity outweighs the benefit for
a testing language.
%\end{comment}

\vspace{-0.1in}
\subsubsection{Less significant features.}
Mini-Haskell provides some smaller conveniences that don't warrant individual
sections but are worth mentioning.
\begin{itemize}
\item \emph{Constructors as functions.}
  As in Haskell, constructors are first class, acting as true functions.  This
  is implemented by simply generating a function that wraps the appropriate
  \texttt{let}-bound \texttt{CON} object in STG.  
%
  Any partially applied
  constructor generates a call to this function.  Fully applied constructors
  generate local \texttt{let}-bound objects to avoid an unnecessary function
  call.

@@

\item \emph{Type annotations.}  The ability to assert the type of some
  variable is often valuable as a sanity-check and provides documentation that
  the compiler can verify.  Mini-Haskell's type annotations (or signatures)
  behave like those in Haskell, being passed to the type checker as % initial
  assumptions.

@@

It is a testament to the talents of the GHC contributors and
maintainers that we have managed to come even this far.  Their
documentation and exposed interfaces have made navigating and
comprehending the internals much more pleasant than expected.

@@

There are several hundred small STG and mini-Haskell programs in the test
suite, which is partitioned into a number of directories which control what
testing is done. Given our goal of exploring various levels of strictness,
most tests are run with all evaluation strategies (three degrees of function
strictness by two degrees of constructor strictness).  Interesting classes of
test programs are those that require less-strict semantics.  For
these programs we test not only that the test leads to the expected result
with less strict semantics, but also that the program returns a
\emph{BLACKHOLE} in the case of a stricter semantics. Students who were
Haskell/STG beginners have contributed substantially to the test suite. We
have also found that the test suite has been of use as a learning tool for
these students to discover what levels of non-strictness are required for their
programs, a benefit that we did not initially envision. Over time we have
developed extensive STG and mini-Haskell preludes of common programming
patterns to aid in the writing of test programs.

Another important class of test programs are those that should fail with a
known error, e.g., a syntax or typing error.  For these programs the test
suite checks the output against an expected error regular expression. 
%This
%class of test programs has proved valuable in development and debugging of
%both the STG and mini-Haskell front-ends.

@@

%We have also implemented a simple logging facility which can be controlled at 
%both compile and run time which helps in determining why a test is failing.
